Article Title

### 2014-06-23 plos.tex
submission candidate

% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}

@@

In the example site (Fig.~\ref{fig:4}), the sparse component had 92.8\% sparsity (or conversely, 7.2\% connectivity: $\mbox{connectivity}=1-\mbox{sparsity}$) with average node degree of 20.9 (\figref{4}{G}). The average node degree, \ie the average number of interactions linking each neuron, is related to connectivity as $\mbox{degree} = \mbox{connectivity}\cdot(p-1)$, where $p$ is the number of neurons. The low-rank component had rank 72, denoting 72 inferred latent units. The number of latent units increased with population size (\figref{5}{A}) but the average number of connections per cell, \ie \emph{node degree} was highly variable (\figref{5}{B}): Several sites, despite their large population sizes, were driven by latent units and had few pairwise interactions. This variability may be explained by differences in brain states and recording quality and warrants further investigation.

@@

\begin{figure}    \floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,center},capbesidewidth=8.3cm}}]{figure}[\FBwidth]
    {\caption{{\bf Properties of $C_{\sf sparse+latent}$ estimates from all imaged sites.} 
    Each point represents an imaged site with its color indicating the population size as shown in panels A and B. The example site from Figures \ref{fig:2} and \ref{fig:4} is circled in blue. 
\\
    {\bf A.} The number of inferred latent units \emph{vs.}~population size.
    {\bf B.} The average node degree for sparse partial correlations vs.~population size.
    {\bf C.} The average sample correlations \emph{vs.}~the average partial correlations (Eq.~\ref{eq:partial}) of the $C_{\sf sparse+latent}$ estimate.
    {\bf D.} The percentage of negative interactions vs.~connectivity in the $C_{\sf sparse+latent}$ estimates.
}
\label{fig:5}}
{\includegraphics{./figures/Figure05.pdf}}
\end{figure}

@@

We examined the relationship of the structure of the $C_{\sf sparse+latent}$ estimates to the differences in orientation preference and to the physical distances separating pairs of cells (Fig.\;\ref{fig:6}).  Five sites with highest connectivitites were included in the analysis. Partial correlations were computed using Eq.~\ref{eq:partial} based on the regularized estimate, including both the sparse and the latent component. Connectivity was computed as the fraction of pairs of cells in each categority connected by non-zero elements (interactions) in the sparse component of the estimate, distinguishing between the positive and negative connectivities.

First, we compared how correlations and connectivity depended on the difference in preferred orientations of pairs of significantly ($\alpha=0.05$) tuned cells. The partial correlations decreased more rapidly with differences in tuning than sample correlations ($p<10^{-9}$ in each of the five sites, two-sample $t$-test on the linear regression coefficients). Positive connectivity decreased rapidly with difference in the preferred orientations ($p<0.005$ in each of the five sites, $t$-test on the logistic regression parameter) whereas negative connectivity did not increase (\figref{6}{D}): The dependence of connectivity on difference in orientation was significantly higher for positive than for negative interactions ($p<0.04$ in each of the five sites, two-sample $t$-test on the logistic regression parameters).

Second, we compared how correlations and connectivity depended on the physical distance separating pairs of cells. We distinguished between lateral displacement, \ie the distance in the \emph{xy} plane between cell pairs located at similar depths $\Delta z < 30 \mu m$, and depth displacement, \ie the difference in depth with lateral displacement under $30\,\mu m$. Again, the partial correlations decreased more rapidly with both lateral and depth displacements ($p<10^{-6}$ in each of the five sites, both lateral and depth, two-sample $t$-test on the linear regression coefficients). 
Lateral positive connectivity decreased rapidly with distance ($p<10^{-6}$ in each of the five sites, $t$-test on the logistic regression parameter) (\figref{6}{E}), so that cells separated laterally by less than 25 $\mu m$ were 3.2 times more likely to be connected than cells separated laterally by more than 150 $\mu m$. The negative connectivity did not significantly change with lateral displacement (\figref{6}{E}): The dependence of connectivity on lateral displacement was significantly higher for positive than for negative interactions ($p<0.05$ in each of the five sites, two-sample $t$-test on the logistic regression parameters). Positive connectivity decreased significantly with increased depth displacement of pairs of cells ($p<0.05$ in each of the five sites, $t$-test on the logistic regression parameter).  Negative connectivity on depth displacement did not reach significance in most sites using the same regression test nor could this dependence be shown to be sufficiently different from the positive connectivity.

\begin{figure}[!ht]
    \begin{center}
        \includegraphics{./figures/Figure06.pdf}
    \end{center}
    \caption{{\bf Dependence of sample correlations and regularized partial correlations on orientation tuning and physical distance.}
    {\bf A--C.} Average partial correlations (red) estimated by $C_{\sf sparse+latent}$ and average sample correlations (black) averaged across multiple imaged sites. At each site the correlations were normalized by the respective average correlation shown in \figref{5}{C}.  The number of sites, $n$, that qualified to be included in the analysis is indicated in each panel. Sites were included if they had at least 20 pairs of neurons in each of the intervals. The error bars indicate the standard error of the mean based on on the number of sites $n$.
    {\bf A.} Average correlations between pairs of neurons tuned to orientation with differences in preferred orientation in the intervals of 0--15$^\circ$, 15--45$^\circ$ and 45--90$^\circ$. 
    {\bf B.} Average correlations between pairs of neurons located at the same depth ($\pm$30$\mu$m) separated by lateral distances in the intervals of 0--25 $\mu$m, 25--75 $\mu$m, 75--150 $\mu$m, and 150+ $\mu$m.
    {\bf C.} Average correlations between pairs of neurons displaced laterally by less than 30 $\mu$m separated in depth by distances in the intervals of 0--25 $\mu$m, 25--60 $\mu$m, and 60+ $\mu$m.
    {\bf D--F.} Normalized connectivity of positive (green) and negative (dark red) interactions from the sparse component obtained from $C_{\sf sparse+latent}$. Normalized connectivity was computed as the fraction of pairs connected by interactions of corresponding signs in each interval divided by the fraction of non-zero interactions across the entire site. }
\label{fig:6}
\end{figure}

@@

In our study, we sought to infer pairwise functional connectivity networks  in cortical microcircuits. We hypothesized that partial correlations correspond more closely to underlying mechanisms than sample correlations when recordings are sufficiently dense.  Since neurons form synaptic connections mostly locally and sparsely \cite{Perin:2011}, we \emph{a priori} favored solutions with sparse partial correlations.  The importance of partial correlations may be justified by the principle of maximum entropy: The maximum entropy distribution on discrete or continuous multivariate domains constrained by the observed mean firing rates, their variances, and correlations is the multivariate normal distribution, where the precision matrix specifies the interactions between cell pairs.  Therefore, under the aforementioned assumptions that the recorded population is sufficiently complete and that the model correctly represents the nature of interactions, the network of partial correlations can be hypothesized to be a better representation of functional dependencies than correlations. %\Kcomment{I am not sure about the previous argument.}

@@

Moreover, this estimation approach provides a graphical representation of the dependencies in the data that can be used to formulate and test hypotheses about the structure of connectivity in the microcircuit. Importantly, the inferred functional interactions were substantially different from the network of the most significant sample correlations \figref{4}{F, H, and I}.  For example, the $C_{\sf sparse+latent}$ estimator reveals a large number of negative interactions that were not present in the sample correlation matrix (\figref{4}{F}) and may reflect inhibitory circuitry.  Although the network of functional interactions inferred by the optimal estimator, $C_{\sf sparse+latent}$, will not be identical to the anatomical connectivity, it may accurately reflect average synaptic connectivity profiles \cite{Song:2005, Oswald:2008, Perin:2011, Levy:2012}. Future experiments utilizing molecular markers for different cell types and follow-up multiple whole-cell \emph{in vitro} experiments \cite{Hofer:2011, Ko:2013} will directly compare the inferred functional connectivity graphs to the underlying anatomical circuitry. Finally, the latent units inferred by the estimator can be analyzed for their physiological interpretation. For example, these latent units may be modulated under different brain states (e.g. slow-wave sleep, attention) and stimulus conditions (e.g. certain types of stimuli may engage feedback connections).

@@

\paragraph{Data processing}
All data were processed in MATLAB using the DataJoint data processing chain toolbox ({\tt datajoint.github.com}) first developed in our lab.

@@

\paragraph{Cross-validation}
To compare the performance of the estimators, we used conventional 10-fold cross validation: Trials were randomly divided into 10 subsets with approximately equal numbers of trials of each condition in each subset. Each subset was then used as the testing dataset with the rest of the data used for estimating the covariance matrix. The resulting validation losses were averaged to produce the \emph{cross-validation loss} reported throughout the paper.

Since each of the regularized estimators had one or two hyperparameters, we used \emph{nested cross-validation}:  The outer loop evaluated the performance of the estimators with the optimal hyperparameter values estimated within the inner loop.  Hyperparameters were optimized by a two-phase search algorithm: random search to find a good starting point for the subsequent pattern search to find the global minimum.  The inner cross-validation loop subdivided the training dataset from the outer loop to perform 10-fold cross-validation in order to evaluate each choice of the hyperparameter values.  Thus the size of the training dataset within the inner loop comprised 81\% of the entire recording.

When the cross-validation loss was not required, only the inner loop of cross-validation was used, applied to the entire dataset.  This approach was used to compute the covariance matrix estimates and their excess loss in the simulation study (\figref{1}{\,Rows 4 and 5}) and to analyze the partial correlation structure of the sparse+latent estimator (Fig.~\ref{fig:4}--\ref{fig:6}).

@@

The MATLAB code for these computations is available at {\tt http://github.com/atlab/cov-est}.

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics{./figures/Supp01.pdf}
\end{center}
\caption{
	{\bf Optimization of hyperparameters of the $C_{\sf sparse+latent}$ estimator.}  
	{\bf A.} The cross-validated loss function (Eq.~\ref{eq:full-loss}) of the data from the example as a function of the values of hyperparameters $\alpha$ and $\beta$ of the minimization problem in Eq.~\ref{eq:ma}. In all panels, the red cross marks the optimal value found by the pattern search algorithm described in Methods.
	{\bf B.} The degree of connectivity ($1-\mbox{sparsity}$) of the sparse component $S$ of the inverse covariance (Eq.~\ref{eq:c-sl}) as a function of $\alpha$ and $\beta$ for the example site.
	{\bf C.} The number of latent units or the rank of the low-rank component $L$ of the inverse covariance (Eq.~\ref{eq:c-sl}) as a function of hyperparameters $\alpha$ and $\beta$ for the example site.
	{\bf D.} The loss function as a function of the the connectivity and the number of latent units for the exapmle site.
}
\end{figure}

>>>>

% amsmath package, useful for mathematical formulas
\usepackage{amsmath}

% amssymb package, useful for mathematical symbols
\usepackage{amssymb}

@@

In the example site (Fig.~\ref{fig:4}), the sparse component had 92.8\% sparsity (or conversely, 7.2\% connectivity: $\mbox{connectivity}=1-\mbox{sparsity}$) with average node degree of 20.9 (\figref{4}{G}). The average node degree, \ie the average number of interactions linking each neuron, is related to connectivity as $\mbox{degree} = \mbox{connectivity}\cdot(p-1)$, where $p$ is the number of neurons. The low-rank component had rank 72, denoting 72 inferred latent units. The number of latent units increased with population size (\figref{5}{A}) but the connectivity was highly variable (\figref{5}{B}): Several sites, despite their large population sizes, were driven by latent units and had few pairwise interactions. This variability may be explained by differences in brain states and recording quality and warrants further investigation.

@@

\begin{figure}    \floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,center},capbesidewidth=8.3cm}}]{figure}[\FBwidth]
    {\caption{{\bf Properties of $C_{\sf sparse+latent}$ estimates from all imaged sites.} 
    Each point represents an imaged site with its color indicating the population size as shown in panels A and B. The example site from Figures \ref{fig:2} and \ref{fig:4} is circled in blue. 
\\
    {\bf A.} The number of inferred latent units \emph{vs.}~population size.
    {\bf B.} The connectivity of the sparse component of partial correlations as a function of population size.
    {\bf C.} The average sample correlations \emph{vs.}~the average partial correlations (Eq.~\ref{eq:partial}) of the $C_{\sf sparse+latent}$ estimate.
    {\bf D.} The percentage of negative interactions vs.~connectivity in the $C_{\sf sparse+latent}$ estimates.
}
\label{fig:5}}
{\includegraphics{./figures/Figure05.pdf}}
\end{figure}

@@

We examined how the structure of the $C_{\sf sparse+latent}$ estimates related to the differences in orientation preference and to the physical distances separating pairs of cells (Fig.\;\ref{fig:6}).  Five sites with highest pairwise connectivities were included in the analysis. Partial correlations were computed using Eq.~\ref{eq:partial} based on the regularized estimate, including both the sparse and the latent component. Connectivity was computed as the fraction of pairs of cells connected by non-zero elements (interactions) in the sparse component of the estimate, distinguishing between the positive and negative connectivities.

First, we analyzed how correlations and connectivity depended on the difference in preferred orientations ($\Delta ori$) of pairs of significantly ($\alpha=0.05$) tuned cells. The partial correlations decayed more rapidly with $\Delta ori$ than did sample correlations ($p<10^{-9}$ in each of the five sites, two-sample $t$-test on the linear regression coefficients). Positive connectivity decreased rapidly with $\Delta ori$ ($p<0.005$ in each of the five sites, $t$-test on the logistic regression parameter) whereas negative connectivity did not decrease (\figref{6}{D}): The slope in the logistic model of connectivity with respect to $\Delta ori$ was significantly higher for positive than for negative interactions ($p<0.04$ in each of the five sites, two-sample $t$-test on the logistic regression parameters).

Second, we compared how correlations and connectivity depended on the physical distance separating pairs of cells. We distinguished between lateral distance, $\Delta x$, in the plane parallel to the pia, and vertical distance, $\Delta z$, orthogonal to the pia.  When considering the dependence on $\Delta x$, the analysis was limited to cell pairs located at the same depth with $\Delta z < 30\,\mu\mbox{m}$; conversely, when considering the dependence on $\Delta z$, only vertically aligned cell pairs with $\Delta x < 30\,\mu\mbox{m}$ were included. Again, the partial correlations decayed more rapidly both laterally and vertically than sample correlations ($p<10^{-6}$ in each of the five sites, for both lateral and vertical distances, two-sample $t$-test on the linear regression coefficients). 
Positive connectivity decayed rapidly with distance ($p<10^{-6}$ in each of the five sites for positive interactions and $p<0.05$ for negative interactions, $t$-test on the logistic regression parameter) (\figref{6}{E}), so that cells separated laterally by less than 25 $\mu\mbox{m}$ were 3.2 times more likely to be connected than cells separated laterally by more than 150 $\mu\mbox{m}$. Although the positive connectivity appeared to decay faster with vertical than with lateral distance, the differences in slopes of the respective logistic regression models was not significant with available data. The negative connectivity decayed significantly slower with distance in the available range (\figref{6}{E and F}): The slope in the respective logistic models with respect to the lateral distance was significantly higher for positive than for negative connectivities ($p<0.05$ in each of the five sites, two-sample $t$-test on the logistic regression parameters).

\begin{figure}[!ht]
    \begin{center}
        \includegraphics{./figures/Figure06.pdf}
    \end{center}
    \caption{{\bf Dependence of sample correlations, regularized partial correlations, and inferred connectivity of the $C_{\sf sparse+latent}$ estimator on the differences in preferred orientations, $\Delta ori$, and physical distances, $\Delta x$ and $\Delta z$.} The error bars mark the standard errors of the means.
    {\bf A--C.} Mean sample correlations (black) and mean partial correlations (red) estimated by $C_{\sf sparse+latent}$ across $n=5$ imaged sites. The values in each bin are normalized by the average across all pairs of their sites.
    {\bf A.} Mean sample correlations and partial correlations of the $C_{\sf sparse+latent}$ estimator between pairs of neurons tuned to orientation as a function of $\Delta ori$.
    {\bf B.} Mean sample correlations and mean partial correlations of the $C_{\sf sparse+latent}$ estimator between horizontally aligned pairs of neurons ($\Delta z<30\,\mu$m) as a function of $\Delta x$.
    {\bf C.} Mean sample correlations and mean partial correlations of the $C_{\sf sparse+latent}$ estimator between vertically aligned pairs of neurons ($\Delta x<30\,\mu$m) as a function of $\Delta z$.
    {\bf D--F.} Mean positive connectivity (green) and mean negative connectivity (dark red) inferred by the $C_{\sf sparse+latent}$ estimator in $n=5$ imaged sites. Positive connectivity in each condition is computed as the fraction of positive partial correlations in the sparse component of the partial correlation matrix meeting the condition normalized by the connectivity in entire site (See \figref{5}{B}). Similarly, negative connectivity is the fraction of negative partial correlations similarly conditioned and normalized.
    {\bf D.} Positive connectivity decreases with $\Delta ori$ whereas negative connectivity does not.
    {\bf E.} Positive connectivity decays with $\Delta x$ whereas negative connectivity does not. 
    {\bf F.} Positive connectivity decays with $\Delta z$. Negative connectivity does not decay for small values of $\Delta z$.
    }
\label{fig:6}
\end{figure}

@@

In this study, we sought to infer pairwise functional connectivity networks  in cortical microcircuits. We hypothesized that partial correlations correspond more closely to underlying mechanisms than sample correlations when recordings are sufficiently dense.  Since neurons form synaptic connections mostly locally and sparsely \cite{Perin:2011}, we \emph{a priori} favored solutions with sparse partial correlations.  The importance of partial correlations may be justified by the principle of maximum entropy: The maximum entropy distribution on discrete or continuous multivariate domains constrained by the observed mean firing rates, their variances, and correlations is the multivariate normal distribution, where the precision matrix specifies the interactions between cell pairs.  Therefore, under the aforementioned assumptions that the recorded population is sufficiently complete and that the model correctly represents the nature of interactions, the network of partial correlations can be hypothesized to be a better representation of functional dependencies than correlations.

@@

Moreover, this estimation approach provides a graphical representation of the dependencies in the data that can be used to formulate and test hypotheses about the structure of connectivity in the microcircuit. Importantly, the inferred functional interactions were substantially different from the network of the most significant sample correlations \figref{4}{F, H, and I}.  For example, the $C_{\sf sparse+latent}$ estimator reveals a large number of negative interactions that were not present in the sample correlation matrix (\figref{4}{F}) and may reflect inhibitory circuitry.

Distances between cells in physical space and in sensory feature space had a stronger effect on the partial correlations estimated by the $C_{\sf sparse+latent}$ estimator than on sample correlations (\figref{6}{A--C}). This difference in decay rates supports the idea that correlations are built up from partial correlations in chains of intermediate cells positioned closer and tuned more similarly to one another, with potentially closer correspondence to anatomical connectivity.  This effect may be at least partially explained by a trivial effect of regularization: $L_1$ penalty employed by the estimator (Eq.~\ref{eq:ma}) suppresses small partial correlations to greater extent than large partial correlations, enhancing the apparent effect of distance and tuning.  These distances also had distinctly different effects on the positive and negative connectivities inferred by the estimator (\figref{6}{D--F}); these differences are not explained by trivial numerical effects and may reflect geometric and graphical features of local excitatory and inhibitory networks. Indeed, the relationships between patterns of positive and negative connectivities inferred by the estimator mirrored some regularities in studies of excitatory and inhibitory synaptic connectivities with respect to distance, cortical layers, and feature tuning \cite{Song:2005, Oswald:2008, Adesnik:2010, Perin:2011, Fino:2011, Hofer:2011, Isaacson:2011, Levy:2012}. For example, while excitatory neurons form synapses within highly specific local cliques \cite{Perin:2011}, inhibitory interneurons form synapses with nearly all excitatory cells within local microcircuits \cite{Fino:2011, Hofer:2011, Packer:2011}.  To further investigate the link between synaptic connectivity and inferred functional connectivity, future experiments will utilize molecular markers for different cell types with follow-up multiple whole-cell \emph{in vitro} experiments \cite{Hofer:2011, Ko:2013} to directly compare the inferred functional connectivity graphs to the underlying anatomical circuitry. Finally, the latent units inferred by the estimator can be analyzed for their physiological interpretation. For example, these latent units may be modulated under different brain states (e.g. slow-wave sleep, attention) and stimulus conditions (e.g. certain types of stimuli may engage feedback connections).

@@

\paragraph{Data processing}
All data were processed in MATLAB using the DataJoint data processing chain toolbox \cite{DataJoint:2014}.

@@

\paragraph{Cross-validation}
To compare the performance of the estimators, we used conventional 10-fold cross validation: Trials were randomly divided into 10 subsets with approximately equal numbers of trials of each condition in each subset. Each subset was then used as the testing sample with the rest of the data used as the training sample for estimating the covariance matrix. The resulting validation losses were averaged to produce the \emph{validation loss} reported throughout the paper.

Since each of the regularized estimators had one or two hyperparameters, we used \emph{nested cross-validation}:  The outer loop evaluated the performance of the estimators with the optimal hyperparameter values estimated within the inner loop.  Hyperparameters were optimized by a two-phase search algorithm: random search to find a good starting point for the subsequent pattern search to find the global minimum.  The inner cross-validation loop subdivided the training dataset from the outer loop to perform 10-fold cross-validation in order to evaluate each choice of the hyperparameter values.  Thus the size of the training dataset within the inner loop comprised 81\% of the entire recording. Fig.~S1 illustrates the dependence of the validation loss on the hyperparameters of the $C_{\sf sparse+latent}$ estimator for the example site shown in Figures \ref{fig:2} and \ref{fig:4} and the optimal value found by the pattern search algorithm.

When the validation loss was not required, only the inner loop of cross-validation was used, applied to the entire dataset.  This approach was used to compute the covariance matrix estimates and their excess loss in the simulation study (\figref{1}{\,Rows 4 and 5}) and to analyze the partial correlation structure of the sparse+latent estimator (Fig.~\ref{fig:4}--\ref{fig:6}) once its superior performance has been established.

@@

The MATLAB code for these computations is available online \cite{Yatsenko:2014c}.

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics{./figures/Supp01.pdf}
\end{center}
\caption{
	{\bf Optimization of hyperparameters of the $C_{\sf sparse+latent}$ estimator.}  
	{\bf A.} The cross-validated loss function (Eq.~\ref{eq:full-loss}) of the data from the example as a function of the values of hyperparameters $\alpha$ and $\beta$ of the minimization problem in Eq.~\ref{eq:ma}. In all panels, the red cross marks the optimal value found by the pattern search algorithm described in Methods.
	{\bf B.} The connectivity ($1-\mbox{sparsity}$) of the sparse component $S$ of the inverse covariance (Eq.~\ref{eq:c-sl}) as a function of $\alpha$ and $\beta$ for the example site.
	{\bf C.} The number of latent units or the rank of the low-rank component $L$ of the inverse covariance (Eq.~\ref{eq:c-sl}) as a function of hyperparameters $\alpha$ and $\beta$ for the example site.
	{\bf D.} The loss function as a function of the the connectivity and the number of latent units for the exapmle site.
}
\end{figure}
